The tracing of trade: exploring the patterns of olive oil production and distribution from Roman Baetica

### 2020-11-02 Paper/jaccardbaetica.tex
review bef submit

%\journal{Journal of Archaeological Science}

@@

\title{The markings of the trade: exploring the patterns of olive oil production in Roman Baetica}

\author[ceipacadress]{Maria Coto-Sarmiento\corref{mycorrespondingauthor}}
\cortext[mycorrespondingauthor]{Corresponding author}
\ead{mcotsar@gmail.com}

\author[edadress]{Xavier Rubio-Campillo}

\address[ceipacadress]{CEIPAC, Department of Prehistory and Archaeology, Montalegre, 6-8, 08001, University of Barcelona, Barcelona, Spain}
\address[edadress]{School of History, Classic \& Archaeology, Room OOM.33, William Robertson Wing, Old Medical School, Teviot Place, University of Edinburgh, UK}

@@

\xavi{reescribir abstract, demasiado general y entonces demasiado específico}
\maria{lo he reeformulado a ver si ahora..}

@@

%Did each province follow a different pattern for its distribution? 
%\xavi{no entiendo esta pregunta} what patterns were followed for the distribution of olive oil to the different provinces? 
%\xavi{no es esta pregunta idéntica a la anterior?} 
%\xavi{esta frase no tiene sentido aquí; o la pones antes o en otro párrafo, pero es raro acabar un set de 2 preguntas con una frase que no sea otra pregunta}

%\xavi{las preguntas no van sobre patrones; los patrones son las trazas en la evidencia que te ayudan a responder a las preguntas}

Advances in the research of the Roman studies have currently led an environment with various interpretations on commercial dynamics \citep{duncan1982economy,
temin_economy_2006,
quantifyingwilson2009}.
%\xavi{more diverse que qué?} 
The application of different quantitative approaches has allowed us to improve our interpretation of the growing amount of archaeological evidence and reveal the dynamics of Roman production %\xavi{en qué sentido? qué quieres decir aquí con complexity? Si no especificas queda muy genérico (rollo "el mundo es muy complejo")}.
\citep{brughmans_roman_2016,
orengo_seeds_2016,bayesian_2018,
coto-sarmiento_identifying_2018,
rubio-campillo_ecology_2018}.

@@

\xavi{me parece que aquí falta más chicha, porque saltas de una cosa super general (complejidad, economía, grandes dinámicas) a unos casos muy específicos. Deberías aquí explicar algo más sobre tu estudio y lo que quieres entender, por ejemplo qué preguntas específicas quieres responder? Como se enlazan con lo que has explicado hasta ahora? Cómo es que ahora saltas a casos de estudio regionales? Las preguntas específicas de los 3 casos de estudio deberían entrelazarse en una pregunta general del paper, que esté vinculada a lo que has explicado hasta ahora}
\maria{he añadido unos párrafos :-)}

@@

The landscape of the \textit{Baetican} province was one of the best regions to face the increased demand for olive oil across the Roman provinces. For this reason, the area saw an increase in productivity as a massive infrastructure of olive oil production was gradually deployed. The production and distribution of olive oil increased from the first to the third centuries AD ~\citep{remesal_concierto}.

@@

This strategic location allowed the transport of olive oil through riverine shipping towards the maritime routes that connected \textit{Baetica} with Mediterranean and Atlantic trade routes towards the rest of the Empire\citep{garcia_vargas_enrique_formal_2010}.

@@

The chronology of the workshops is wide spanning from the first to the third centuries AD
%\xavi{qué quiere decir que la cronología es diversa?} 
\citep{millet_anforas_1998,rodriguez_baetican_1998,chic2005comercio}. 
This fact is shown by the archaeological evidence that displays a highly specialised production with a long activity \citep{remesal_anforas_2004}. Thus, the specialisation of this production made amphorae with apparently few changes in their forms, reaching a high degree of standardisation.

\xavi{deberías desarrollar esto un poco más; qué tipo de evidencia demuestra esto? pocos cambios en qué? los talleres? las tipologías anfóricas?}
\maria{hecho}

The importance of this olive oil trade is revealed by the massive amount of Dressel 20 type amphorae found across the Roman provinces \citep{dressel_ricerche_1878,
millet_anforas_1998}. This amphora has been commonly associated with the transportation of Baetican olive oil for supplying military camps and civil settlements \citep{berni_millet_epigrafianforica_2008}.%\xavi{me parece repetitivo, porque unos párrafos antes ya hablabas de lo mismo. Quizás unificar texto aquí o allá?}.

@@

\xavi{esto ya lo has mencionado; quizás podrías explicar en 2 líneas la capacidad y forma de la dressel 20}.
\maria{hecho y borrado párrafo}

@@

Dressel 20 was the amphora most stamped during the Roman Empire ~\citep[18]{millet_anforas_1998}.

%The usage period is associated with a large amphora production for almost three centuries\xavi{esto está repetido no?}.

The large sample size of recovered Dressel 20 stamps is distributed across all the western provinces of the Roman Empire. For this reason, there are several publications of this type of evidence discussing its origin, its meaning, its long period of use and its spatial distribution~\citep{dressel_ricerche_1878,
rodriguez_economioleicola_1977,
chicepi1985,millet_anforas_1998, remesal_sellar_2016} (Fig.\ref{amphora}).

\xavi{si dices que es muy popular debería haber aquí más citas aparte de Remesal}
\maria{he incluido nuevas citas}

@@

Most stamps display a large activity of production that can be difficult to date using an accurate chronology. They have been frequently dated with the consular dating by studying \textit{tituli picti} found in \textit{Monte Testaccio} \citep{Testaccio1, berni_millet_epigrafianforica_2008}.
Despite \textit{tituli picti} contain important information about chronology, they have been still found in few quantities mostly concentrated in an only settlement such as \textit{Monte Testaccio}. In addition, the chronology of stamps could be also biased taking into account that amphorae were deposited after being produced and consumed\xavi{esto necesita un poco de desarrollo. quieres decir que el código se encuentra en contextos de cronologías diferentes no? Y lo del Testaccio, pos eso, que hay pocos y todos de un yacimioento, lo que añade un bias}. 
\maria{desarrollado}

The use of amphoric stamps can be a good proxy to explore the system of olive oil production and trade developed in \textit{Baetica} province. These stamps are typically found on handles but they were also imprinted on rims and the amphora body~\citep{millet_anforas_1998}. The content information of the stamps is shown in different forms and letter content and it seems that there was not a unique criterion.
%\xavi{qué information? El código de letras?}. 
Stamps often displayed a code of three letters and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}. The three letters correspond to initial forms of a name but they can also appear complete. 
\xavi{que quieres decir con abbreviated o complete? Como abrevias un código de 3 letras?}
\maria{corregido}

@@

%Additionally, a large percentage of Dressel 20 containers were not stamped thus complicating the interpretation~\xavi{esta última frase aquí es un poco random no?}.

@@

\xavi{acabas de decir que tampoco se sabe qué es el sello, así que esta frase queda rara} 
\maria{he cambiado la frase de sitio para que se contextualice mejor}

@@

%Neither the use of written records have allowed providing enough information that can explain the economic role of \textit{Baetica} province in the Roman organisation.\xavi{esta frase esta fuera de contexto. Si la pones entonces deberías expandir la idea de frase a párrafo, y poner referencias}.
%\maria{he eliminado la última frase}

@@

The creation of new provinces required the Roman Empire the arrival of resources through Mediterranean and Atlantic routes.
%\xavi{no entiendo esto...allowed o required?}. 
This led to a gradual change in the economic and social structures the trade networks that supplied the Roman army became more complex. Augustus's administration created the figure of the \textit{praefectura annonae} to organize wheat supply. The role of the \textit{praefectura annonae} was mainly focused on providing through \textit{frumentationes} a fixed monthly amount of wheat to each Roman citizen~\citep{remesal_annona_1986,remesal_concierto}\xavi{pero estabas hablando del ejército no? Has hecho un salto aquí a los ciudadanos}. 
\maria{Pero porque la annona estaba enfocada a los ciudadanos romanos y luego hablo después que hay otras hipótesis que dicen que iba más allá a los militares}

@@

\xavi{pero no hay evidencia? Las fuentes escritas no lo mencionan?}
\maria{He cambiado authors por hipotesis porque me quería referir a investigadores}.

@@

Olive oil supply seems to be particularly intense in militarised provinces that functioned as the borders of the Empire such as \textit{Britannia} and \textit{Germania}. The high percentage of Dressel 20 amphorae found in sites nearby military garrisons suggests the existence of diverse supply dynamics as there were trade processes flowing from military sites to civil settlements~\citep{remesal_annona_1986, carreras_britannia_1998}
\xavi{lo que quieres decir es que tienes muchas ánforas en sitios civiles cerca de militares? no queda claro}.
\maria{sí, pero creo que yo no he escrito está frase xD}

An explanation for the interest to export olive oil for the military garrison in the Roman provinces is still unclear. The origin of this exportation may be related to two aspects: a) cultural consumption whose olive oil is used by cultural reasons such as group identification or by habit and b) economical and political reasons to create a redistributive system mechanism to supply military camps. 
\maria{esta frase está explicada mejor}

%may be related to two aspects according with some hypotheses: a) cultural consumption whose product is consumed by cultural reasons such as to identity or habit and b) economical reasons where olive oil distribution is based on the transport costs, unlike other products \citep[69-70]{carreras_britannia_1998}\xavi{no pillo esta frase}..

@@

\xavi{faltaría un enlace a la siguiente sección, queda muy abrupto}
\maria{he añadido varios párrafos a ver si se enlaza}

@@

The evidence for olive oil consumption in \textit{Britannia} is scarce before the Roman conquest \citep{funari_corpus_1996,carreras_abastecimiento_2003}. The local population did not consume this product and it was not produced on the region as the landscape and climate of the British Isles was not suitable for olive oil trees\citep[161]{monfort_britanniaen_1998}. Based on the archaeological records, this absence of olive oil changed with the arrival of a large amount of Dressel 20 amphorae to \textit{Britannia} from the first century A.D \citep{peacock_amphorae_1991,
carreras_britannia_1998}. At this moment, we detect an increase of the olive oil exportation concurring with the displacement of legions during the military campaigns \citep[161]{monfort_britanniaen_1998}
%\xavi{en qué siglo?}

@@

The presence of Dressel 20 stamps in military camps in \textit{Britannia} has been widely studied in Roman archaeology \citep{williams_importation_1983,
funari_corpus_1996,
carreras_britannia_1998,carreras_abastecimiento_2003}.\xavi{si dices widely studied no puedes referencias solo 1 autor}
\maria{he añadido más autores}

@@

%such as \textit{Germania} (Remesal Rodrıguez, 1986)\xavi{por qué such as Germania? Aún no has explicado qué hacen en Germania}.

@@

\xavi{pero entonces la dressel 23 de donde viene? Cuidao que estás diciendo que se acaban las dressel 20, pero esto implica que hay menos aceite de oliva o que viene de otras partes o que va en recipientes distintos a la dressel 20?}
\maria{he eliminado esta frase para que quede más claro}

@@

%Julius Caesar's campaigns in \textit{Gallia} during 1st BC opened the door to the invasion of \textit{Germania} during Augustus rule~\citep{remesal_annona_1986,remesal_baetica_2002ç
\xavi{no hay un resumen de la conquista similar para Britannia, queda raro}. 
\maria{lo he eliminado}

Roman studies exploring the presence of Dressel 20 amphorae in the provinces Inferior and Superior during Roman Empire have not had the same impact as the rest of the provinces \citep[293]{remesal_baetica_2002}.
%\xavi{no entiendo esta frase. Que qué provincia? Germania son 2 provincias no?}

The massive importation of Dressel 20 in both \textit{Germania} provinces could be explained due to the archaeological sources found mostly in German \textit{limes}. 
By contrast, it is still unknown if trade agents participated in the distribution of Baetican olive oil in \textit{Germania} or the supply was exclusively focused on the Roman army garrisons\citep[156]{remesal_germn_2010}
\xavi{quieres decir que aquí sólo encontramos dressel 20 en yacimientos militares porque no se han excavado muchos civiles?}.
\maria{lo he mejorado, a ver si se entiende}

@@

%\xavi{cita?}

@@

\xavi{pero justo antes decías que no había en civiles no?}. 
\maria{yo no he dicho eso :-[}

It seems that some Baetican centres could have been assigned to a specific province for the olive oil supply to military garrisons \citep[125]{remesal_concierto}. 
\xavi{the support of olive oil no se entiende. Además, falta alguna cita y desarrollar esta idea}. 
\maria{mejorado, creo}
However, this hypothesis is hard to assess given the current lack of written evidence related to the meaning of the amphoric stamps. 
\xavi{related to BLABLABLA. Habría que detallar un poco porque hay un montón de evidencia arqueológica del limes, así que qué es exactamente lo que falta?}. 
\maria{mejorado,creo}

Previous studies suggested that the supply to the German limes was mainly based on riverine transport, but recent works suggest that the maritime route through the Atlantic Ocean could have been more important than expected (both for \textit{Germania} and \textit{Britannia}).~\citep{remesal_germn_2010,rubio-campillo_ecology_2018}.
%\xavi{lo del océano atlántico está un poco descontextualizado; no sería mejor ponerlo al final de esta sección?}

@@

%To do that, we use the CEIPAC database to collect stamps from different places. The CEIPAC dataset contains over 50.000 of epigraphy records found in amphorae, mostly from \textit{Monte Testaccio}.

@@

%\xavi{estaría bien ampliar un poco esto; qué organizaban exactamente?}

@@

%It is important to mention that some workshops exhibited the same coordinates but they were catalogued as different workshops in Roman Studies (i.e. nearby amphora workshops in \textit{Baetica} named as Tesorillo de Doña Mencia and Doña Mencia). The analysis took into account this particularity but some biases can occur due to the archaeological dataset.

\xavi{No se entiende; deberías explicarlo un poco mejor y citar algo si dices eso. Piensa que la mayoría de lectores no sabrán quién es Doña Mencia}. 
\maria{he reformulado la frase a ver si ahora se entiende}

@@

%\xavi{el gráfico es dificil de entender...yo leo que hay unos 20 workshops con 1 solo stamp, por lo que entiendo que no es cierto que la mayoría solo tengan 1 sello (hay como 50 workshops con entre 2 y 30 sellos no?)} 
\maria{he cambiado la frase a ver si se entiende mejor}

The type of distribution is also frequent in amphora production where we observed a self-organised complex system pattern with a major concentration of the number of stamps in a few workshops \citep{bayesian_2018,coto-sarmiento_identifying_2018}\xavi{no lo veo; recuerda que para el rollo bayesiano tenías una línea recta al hacedr el log de los 2 ejes, no solo 1, y este plot no está construído así no?}.
\maria{son dos ejes, no? ¿aquí no se ve línea recta? Lo mismo si lo hago de puntitos...}

@@

%\xavi{una cosa, aquí has pasado de sellos a códigos, y los valores son distintos que en el gráfico anterior. Deberías clarificarlo antes de empezar a hablar de la figura}.

According to previous studies, those workshops became the most important centres of amphora production in the region \citep{rodriguez_economioleicola_1977,
arva_1997}.
\xavi{cita de estos estudios}
\maria{citados}

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/frequency}
\caption{Distribution of the number of different code stamps (X axis) for each \textit{conventus} (Y axis). Each dot corresponds to a workshop sorted by different areas. Colours are represented by areas divided into Hispalis (red), Astigi (green) and Corduba (blue)}
\label{frequency}
\end{figure}

@@

\xavi{pero esto no se hizo en el de Baetica no? Por qué?}
\maria{lo he especificado}

@@

%We analysed a dataset of 2219 stamps from different centres in \textit{Britannia}. 
%CITAR (Callender, 1965; Carreras Monfort y Funari,1998; Ayllón-Martı́ et al., 2018). 
In \textit{Britannia}, we studied a total sample of 1765 stamps from 46 sites displaying 968 unique codes.

\xavi{pero esto no concuerda con las 2219 de antes...es por descartar los inciertos? Hay que dejarlo claro}
\maria{clarificado}

The centres located in \textit{Britannia} can be seen in Figure~\ref{britannia}.

@@

The approach proposed here aimed at exploring links between production and consumption centers through the identification of common amphorae codes found in different sites. For this reason we measured the similarity between amphora workshops and/or consumption sites by quantifying a pairwise distance or dissimilarity index between places (i.e. to what extent the stamp codes found on these sites were different). The chosen dissimilarity measure was the Morisita-Horn index \citep{morisita_measuring_1959, horn_measurement_1966}. This method was applied to measure the dissimilarity between different samples of sets. Generally, it describes the dissimilarity between the system of two communities based on the idea of inverse correlation between diversity and species \citep{magurran_why_1988}. In our case, stamps from different sites will be used as a proxy to measure their dissimilarity by analyzing their code. 
\xavi{citar reforzar más esta idea}.
\maria{he añadido unas líneas a ver si están bien o no xD}

@@

Considering our dataset as a non-uniform sample, this method provides a useful tool to handle large samples with different sizes and diversity \citep{wolda_similarity_1981}. Morisita-Horn index gives a value between 0 (sites have exact stamp codes and frequency distribution) and 1 (complete difference between stamp codes). To apply Morisita-Horn we first calculated the number of times that each stamp code appeared in an amphora workshop. 
This method allowed us to bear in mind the number of frequency of stamps for each workshop. 
%This method allowed us to bear in mind a similar number of times for each repeated stamp per workshop\xavi{no entiendo esta frase}.

@@

%By contrast, most sites did not show a strong similarity in stamp correlation\xavi{qué quieres decir? porque lo de que hay una similaridad baja ya lo has repetido arriba}. We also did not find any groups of similar stamps in a specific place\xavi{tampoc entiendo esto. que no hay sellos similares en un solo sitio...?}.

@@

A higher concentration of sites sharing similar stamps was found in areas eminently militarised and close to German \textit{limes}, although most sites mostly showed different stamps. 
\maria{he cambiado esta frase}
%even if most sites mostly showed different stamps~\xavi{quieres decir que los sitios más cercanos al limes son más similares? no queda claro y hay que detallarlo un poco más}.

%\xavi{esta frase me la petaría; no aporta demasiado: The interpretation of the results suggests that there is still no possibility to determine the existence of a defined pattern that reflects in more detail the route of this production in the area of the German limes.}

@@

%The general purpose was to explore if such differences found in the stamp codes could play an important role in the Roman market\xavi{este párrafo es un poco reiterativo no?}.

@@

%In general, the majority of stamps were located in different amphora workshops\xavi{estas 2 frases repiten cosas dichas antes}. 
\maria{he quitado la última frase pero la anterior no se repite}

@@

%These stamp codes detected in closer workshops do not move from other distant workshops. In other words, the stamps tend to remain in the same area and different stamps were located in a same amphora workshop\xavi{de nuevo estas 2 frases repiten cosas de antes}.

This unique link between workshop and stamp code can be interpreted in different ways. First, it is possible that each workshop operated independently and did not normally collaborate with other workshops.  Second, stamp similarity in closer workshops could be linked to a spatial pattern. It is more probable than closer workshops tend to share more traits than distant workshops. While the role of the river was significant for the distribution of amphorae, river connectivity amongst workshops does not seem to show relevancy for the distribution or concentration of stamps in a specific river area.
\xavi{qué quieres decir con esto? que los diferentes ríos no muestran ningún patrón?}
\maria{he añadido cosas a la frase para ver si se entiende}

Alternatively, the presence of different stamps in the same workshop could imply some kind of organisation from within the workshop only affecting nearby centres \citep{juanmorostesis}. The stamped amphorae could have allowed potters to organise the different production batches for posterior commercialisation.
%\xavi{estos 2 párrafos igual deberían ir a la discusión no? La conclusión debería ser algo más general y aquí estás aún interpretando resultados}.

@@

A combined exploration of the map and the dendrogram also suggests an interesting pattern: centres with higher similarity values are closer to the coast (North Sea and Celtic Sea). This may indicate that the Atlantic route could have played an essential role in transporting olive oil to the area of \textit{Britannia}, since in the places where there is greater similarity they are found in different strategic points near the sea
\xavi{por qué más conectividad implica más similaridad? no será porque encuentras más ánforas?}
\maria{no entiendo este comentario :-(}.

@@

The areas mostly militarised share similar stamps than civil areas. However, we do not detect a concrete pattern regarding the distribution of the stamps in the German limes \citep{xanten2018}. This could be also interpreted by an intensity bias where military centres have been mostly excavated than civil centres

@@

\xavi{esto no iría en la discusión? Por otra parte, no entiendo qué quieres decir exactamente, igual hay que desarrollarlo}.
\maria{lo he incluido aquí y lo he cambiado a ver si se entiende}
%Rather, it would imply that olive oil production was distributed from non-specific production centres.

@@

To conclude, the method presented here provides a useful framework to improve our understanding of the organisation and processes that allowed the Roman Empire to manage a massive and highly specialized production of essential goods. This work has identified similarities between amphoric stamps and spatial distribution in the case of the amphoric production within the Roman Empire
%\xavi{qué diferencias?}. 
The growing amount of archaeological data requires of large-scale quantitative methods such as the one presented here to identify and explore complex patterns, thus allowing us to infer dynamics and improve our interpretation of complex economic processes of past societies.

>>>>

\journal{Journal of Archaeological Method and Theory}

@@

%\title{The markings of the trade: exploring the patterns of olive oil production in Roman Baetica}
\title{The tracing of trade: exploring the patterns of olive oil production and distribution from Roman Baetica}

\author[ceipacadress]{María Coto-Sarmiento\corref{mycorrespondingauthor}}
\cortext[mycorrespondingauthor]{Corresponding author}
\ead{mcotsar@gmail.com}

\author[edadress,ubadress]{Xavier Rubio-Campillo}

\address[ceipacadress]{Department of Prehistory and Archaeology, Montalegre, 6-8, 08001, University of Barcelona, Barcelona, Spain}
\address[edadress]{School of History, Classic \& Archaeology, Room OOM.33, William Robertson Wing, Old Medical School, Teviot Place, University of Edinburgh, UK}
\address[ubadress]{DIDPATRI, Passeig de la Vall d'Hebrón, 171, Barcelona, Spain}

@@

@@

Advances in the research of the Roman studies have currently led an environment with various interpretations on commercial dynamics \citep{duncan1982economy,
temin_economy_2006,
quantifyingwilson2009}.
%\xavi{more diverse que qué?} 
The application of different quantitative approaches has allowed us to improve our interpretation of the growing amount of archaeological evidence and reveal the dynamics of Roman production
\citep{brughmans_roman_2016,
orengo_seeds_2016,bayesian_2018,
coto-sarmiento_identifying_2018,
rubio-campillo_ecology_2018}.

@@

@@

The landscape of the \textit{Baetican} province was one of the best regions to face the increased demand for olive oil across the Roman provinces. For this reason, the area saw an increase in productivity as a massive infrastructure of olive oil production was gradually deployed. The production and distribution of olive oil increased from the first to the third centuries AD~\citep{remesal_concierto}.

@@

This strategic location allowed the transport of olive oil through riverine shipping towards the maritime routes that connected \textit{Baetica} with Mediterranean and Atlantic trade routes towards the rest of the Empire~\citep{garcia_vargas_enrique_formal_2010}.

@@

The chronology of the workshops is wide spanning from the first to the third centuries AD\citep{millet_anforas_1998,rodriguez_baetican_1998,chic2005comercio}. 
This fact is shown by the archaeological evidence that displays a highly specialised production with a long activity \citep{remesal_anforas_2004}. Thus, the specialisation of this production made amphorae with apparently few changes in their forms, reaching a high degree of standardisation.

The importance of this olive oil trade is revealed by the massive amount of Dressel 20 type amphorae found across the Roman provinces \citep{dressel_ricerche_1878,
millet_anforas_1998}. This amphora has been commonly associated with the transportation of Baetican olive oil for supplying military camps and civil settlements \citep{berni_millet_epigrafianforica_2008}.

@@

@@

Dressel 20 was the amphora most stamped during the Roman Empire ~\citep[18]{millet_anforas_1998}. The large sample size of recovered Dressel 20 stamps is distributed across all the western provinces of the Roman Empire. For this reason, there are several publications of this type of evidence discussing its origin, its meaning, its long period of use and its spatial distribution~\citep{dressel_ricerche_1878,
rodriguez_economioleicola_1977,
chicepi1985,millet_anforas_1998, remesal_sellar_2016} (Fig.\ref{amphora}).

@@

Most stamps display a large activity of production that can be difficult to date using an accurate chronology. They have been frequently dated with the consular dating by studying \textit{tituli picti} found in \textit{Monte Testaccio} \citep{Testaccio1, berni_millet_epigrafianforica_2008}.
Despite \textit{tituli picti} contain important information about chronology, they have been still found in few quantities mostly concentrated in an only settlement such as \textit{Monte Testaccio}. In addition, the chronology of stamps could be also biased taking into account that amphorae were deposited after being produced and consumed.

The use of amphoric stamps can be a good proxy to explore the system of olive oil production and trade developed in \textit{Baetica} province. These stamps are typically found on handles but they were also imprinted on rims and the amphora body~\citep{millet_anforas_1998}. The content information of the stamps is shown in different forms and letter content and it seems that there was not a unique criterion.

Stamps often displayed a code of three letters and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}. The three letters correspond to initial forms of a name but they can also appear complete.

@@

%Additionally, a large percentage of Dressel 20 containers were not stamped thus complicating the interpretation

@@

@@

@@

The creation of new provinces required the Roman Empire the arrival of resources through Mediterranean and Atlantic routes.

This led to a gradual change in the economic and social structures the trade networks that supplied the Roman army became more complex. Augustus's administration created the figure of the \textit{praefectura annonae} to organize wheat supply. The role of the \textit{praefectura annonae} was mainly focused on providing through \textit{frumentationes} a fixed monthly amount of wheat to each Roman citizen~\citep{remesal_annona_1986,remesal_concierto}

@@

@@

Olive oil supply seems to be particularly intense in militarised provinces that functioned as the borders of the Empire such as \textit{Britannia} and \textit{Germania}. The high percentage of Dressel 20 amphorae found in sites nearby military garrisons suggests the existence of diverse supply dynamics as there were trade processes flowing from military sites to civil settlements~\citep{remesal_annona_1986, carreras_britannia_1998}.

An explanation for the interest to export olive oil for the military garrison in the Roman provinces is still unclear. The origin of this exportation may be related to two aspects: a) cultural consumption whose olive oil is used by cultural reasons such as group identification or by habit and b) economical and political reasons to create a redistributive system mechanism to supply military camps \citep[69-70]{carreras_britannia_1998}.

@@

@@

The evidence for olive oil consumption in \textit{Britannia} is scarce before the Roman conquest \citep{funari_corpus_1996,carreras_abastecimiento_2003}. The local population did not consume this product and it was not produced on the region as the landscape and climate of the British Isles was not suitable for olive oil trees\citep[161]{monfort_britanniaen_1998}. Based on the archaeological records, this absence of olive oil changed with the arrival of a large amount of Dressel 20 amphorae to \textit{Britannia} from the first century A.D \citep{peacock_amphorae_1991,
carreras_britannia_1998}. At this moment, we detect an increase of the olive oil exportation concurring with the displacement of legions during the military campaigns \citep[161]{monfort_britanniaen_1998}.

@@

The presence of Dressel 20 stamps in military camps in \textit{Britannia} has been widely studied in Roman archaeology \citep{williams_importation_1983,
funari_corpus_1996,
carreras_britannia_1998,carreras_abastecimiento_2003}.

@@

@@

@@

Roman studies exploring the presence of Dressel 20 amphorae in the provinces Inferior and Superior during Roman Empire have not had the same impact as the rest of the provinces \citep[293]{remesal_baetica_2002}.

The massive importation of Dressel 20 in both \textit{Germania} provinces could be explained due to the archaeological sources found mostly in German \textit{limes}. 
By contrast, it is still unknown if trade agents participated in the distribution of Baetican olive oil in \textit{Germania} or the supply was exclusively focused on the Roman army garrisons\citep[156]{remesal_germn_2010}.

@@

@@

It seems that some Baetican centres could have been assigned to a specific province for the olive oil supply to military garrisons \citep[125]{remesal_concierto}.

However, this hypothesis is hard to assess given the current lack of written evidence related to the meaning of the amphoric stamps.

Previous studies suggested that the supply to the German limes was mainly based on riverine transport, but recent works suggest that the maritime route through the Atlantic Ocean could have been more important than expected (both for \textit{Germania} and \textit{Britannia}).~\citep{remesal_germn_2010,rubio-campillo_ecology_2018}.

@@

@@

@@

@@

The type of distribution is also frequent in amphora production where we observed a major concentration of the number of stamps in few workshops\citep{bayesian_2018,coto-sarmiento_identifying_2018}.

%\xavi{no lo veo; recuerda que para el rollo bayesiano tenías una línea recta al hacedr el log de los 2 ejes, no solo 1, y este plot no está construído así no?}.
%\maria{son dos ejes, no? ¿aquí no se ve línea recta? Lo mismo si lo hago de puntitos...}

@@

According to previous studies, those workshops became the most important centres of amphora production in the region \citep{rodriguez_economioleicola_1977,
arva_1997}.

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/frequency}
\caption{Distribution of the number of different code stamps (X axis) for each \textit{conventus} (Y axis). Each dot corresponds to a workshop sorted by different areas. Colours are represented by areas divided into \textit{Hispalis} (red), \textit{Astigi} (green) and \textit{Corduba} (blue)}
\label{frequency}
\end{figure}

@@

@@

%We analysed a dataset of 2219 stamps from different centres in \textit{Britannia}. 
%CITAR (Callender, 1965; Carreras Monfort y Funari,1998; Ayllón-Martı́ et al., 2018). 
In \textit{Britannia}, we studied a total sample of 1765 stamps from 46 sites displaying 968 unique codes. The centres located in \textit{Britannia} can be seen in Figure~\ref{britannia}.

@@

The approach proposed here aimed at exploring links between production and consumption centres through the identification of common amphorae codes found in different sites. For this reason we measured the similarity between amphora workshops and/or consumption sites by quantifying a pairwise distance or dissimilarity index between places (i.e. to what extent the stamp codes found on these sites were different). The chosen dissimilarity measure was the Morisita-Horn index \citep{morisita_measuring_1959, horn_measurement_1966}. This method was applied to measure the dissimilarity between different samples of sets. Generally, it describes the dissimilarity between the system of two communities based on the idea of inverse correlation between diversity and species \citep{magurran_why_1988}. In our case, stamps from different sites will be used as a proxy to measure their dissimilarity by analyzing their code.

@@

Considering our dataset as a non-uniform sample, this method provides a useful tool to handle large samples with different sizes and diversity \citep{wolda_similarity_1981}. Morisita-Horn index gives a value between 0 (sites have exact stamp codes and frequency distribution) and 1 (complete difference between stamp codes). To apply Morisita-Horn we first calculated the number of times that each stamp code appeared in an amphora workshop. 
This method allowed us to bear in mind the number of frequency of stamps for each workshop.

@@

@@

A higher concentration of sites sharing similar stamps was found in areas eminently militarised and close to German \textit{limes}, although most sites mostly showed different stamps.

@@

@@

@@

This unique link between workshop and stamp code can be interpreted in different ways. First, it is possible that each workshop operated independently and did not normally collaborate with other workshops.  Second, stamp similarity in closer workshops could be linked to a spatial pattern. It is more probable than closer workshops tend to share more traits than distant workshops. While the role of the river was significant for the distribution of amphorae, river connectivity amongst workshops does not seem to show relevancy for the distribution or concentration of stamps in a specific river area.

Alternatively, the presence of different stamps in the same workshop could imply some kind of organisation from within the workshop only affecting nearby centres \citep{juanmorostesis}. The stamped amphorae could have allowed potters to organise the different production batches for posterior commercialisation.

@@

A combined exploration of the map and the dendrogram also suggests an interesting pattern: centres with higher similarity values are closer to the coast (North Sea and Celtic Sea). This may indicate that the Atlantic route could have played an essential role in transporting olive oil to the area of \textit{Britannia}, since in the places where there is greater similarity they are found in different strategic points near the sea.

@@

The areas mostly militarised share similar stamps than civil areas. However, we do not detect a concrete pattern regarding the distribution of the stamps in the German limes \citep{xanten2018}. This could be also interpreted by an intensity bias where military centres have been mostly excavated than civil centres.

@@

@@

To conclude, the method presented here provides a useful framework to improve our understanding of the organisation and processes that allowed the Roman Empire to manage a massive and highly specialized production of essential goods. This work has identified similarities between amphoric stamps and spatial distribution in the case of the amphoric production within the Roman Empire.

The growing amount of archaeological data requires of large-scale quantitative methods such as the one presented here to identify and explore complex patterns, thus allowing us to infer dynamics and improve our interpretation of complex economic processes of past societies.
